Effect of Sociological Factors on Visual Statistical Inference

### 2024-12-30 wiley_paper/wiley_paper.Rnw
Add files via upload

The lineup protocol introduced in \citet{buja:2009} can be used to quantify the significance of graphical findings in the exploratory data analysis process. This methodology is  part of what is called visual statistical inference, which has been developed further  and validated by \citet{majumder:2013} in simulation studies of head to head comparisons with conventional inference. One of their major contributions   is to define the power of  visual tests and to provide a method for obtaining the power for a particular lineup. In some scenarios, the power of a visual test  was seen to outperform that of a conventional test.

In visual inference, the test statistic is a plot of the observed data, called the {\it (actual) data plot}. To create a lineup the data plot is placed randomly in a grid of null plots. Null plots  are rendered from data generated as  specified by a null hypothesis. Often, this null hypothesis assumes independence or no structure in the data. An observer is then asked to evaluate the lineup. If the  actual data plot is detected by the observer, this counts as evidence against the null hypothesis. Based on the number of evaluations by independent observers and the number of times that the data plot is identified, we are  able to evaluate the significance of the graphical relationship in the same way and the same rigor as a conventional hypothesis test.

\begin{figure}[hbtp] 
   \centering
   \includegraphics[width=0.99\textwidth]{electoral-5-13.pdf} 
   \caption{ \label{fig:elect} Which one of the plots, labelled 1 through 20 from top left to bottom right, is the most different from the others?}
\end{figure}

Figure~\ref{fig:elect} displays an example lineup, one of the plots is based on the observed data, while the remaining 19 plots show  data generated from a null model. Which one of the 20 plots is the most different from the others? 
When asked this question, 12 out of 72 observers picked the  data plot located in panel number $4^2-3$ in the lineup\footnote{The little piece of calculus used for describing the location of the data plot in the lineup provides a small cognitive hurdle that is supposed to enable the interested reader to inspect the lineup without being biased by knowing the location of the data plot.}.  The corresponding $p$-value is 0.0077, indicating sufficient evidence to reject the null hypothesis.

How do we interpret this finding, though? For that, we need to know the context of the data and we need to have more information about how the null plots were generated. 
This particular example investigates the results from the 2012 US presidential election in comparison to  poll results just prior to the election. 
Although this example is more simplistic than most of the tests conducted to date, it serves the purpose of illustrating the lineup protocol. The data consists of the difference in poll results between the two (major) presidential candidates, Obama and Romney, for all states. Each panel in Figure~\ref{fig:elect} shows an `electoral building' \citep{mosley2010} where each state in the union is represented by a rectangle. The difference in poll results is plotted horizontally, while the height of each box corresponds to the number of that state's electoral votes. Color indicates party affiliation.  The null hypothesis is that ``the election results are consistent with the polls''. Therefore the polling results provide the null model from which data is simulated. A normal model with mean and standard deviation based on a poll and its margin of error is used to simulate different scenarios that might have resulted on election day, if the polls were on target. Each null data set is generated as a set of draws from this model. % normal distribution, with mean equal to the difference in poll percentage of the latest state poll results, and standard deviation equal to 2.5, approximating a margin of error of 5\%. 
These samples are plotted as electoral buildings, and the plot based on the actual election results is placed randomly among the null plots in a lineup of size $5\times 4$. If the null hypothesis is true, the  data plot should look  like any of the other plots, and not be identifiable by an observer. Figure~\ref{fig:tower} shows the plot of the electoral building with the additional context information and labels.

\begin{figure}[hbtp] 
   \centering
   \begin{minipage}[c]{0.49\textwidth}
   \includegraphics[width=\textwidth]{tower.pdf}
   \end{minipage}
    \hfill
   \begin{minipage}[c]{0.45\textwidth}
   Poll aggregator 1 \\
      \includegraphics[width=\textwidth]{freedom-sim.pdf} 
   Poll aggregator 2\\
      \includegraphics[width=\textwidth]{rcp-sim.pdf} 
   \end{minipage}
   \caption{ \label{fig:tower} Electoral building plot of the results of the 2012 U.S. Presidential Election (left). On the right two histograms showing simulated poll results for the Democratic presidential candidate using  polling averages collected from two different sources, Freedom's Lighthouse Averages (top), RealClearPolitics (bottom).  If the polls matched the electoral result the simulated results would be centered at the electoral result. This happens for the RealClearPolitics polls (bottom) but not for Freedom's Lighthouse.  For the latter, only 5.33\% of simulated results were as high as the actual election result, which converts to a $p$-value of 0.0533, indicating an almost statistically significant difference. It hints at the polls from this source under-estimating the vote for Obama. This source was used to generate the null plots in the lineup in Figure~\ref{fig:elect}. The $p$-value from the lineup protocol, was 0.0077, which more strongly concludes that these polls mismatched the actual result.}
\end{figure}
The lineup of Figure~\ref{fig:elect}, as well as the other lineups in the manuscript are created using the nullabor package by \citep{nullabor}.

A lineup can be evaluated by a single person or multiple observers. A distribution similar to the binomial distribution, but adjusted for dependencies introduced by the lineup scenario, is used to calculate the $p$-value based on the number of times observers identify the actual data plot, which provides the information needed to make a decision on rejecting or failing to reject the null hypothesis \citep{vinference}.  To avoid expectation errors \citep{meilgaard}, particularly in emotionally charged areas such as election results, observers should not be aware of the data that constitutes a lineup, and should therefore not have seen the actual data plot before inspecting the lineup. This is the reason that we revealed the context of the election problem only after the interested reader had a chance to evaluate the lineup.

When presenting a lineup to an observer the accompanying question should be phrased in as general a manner as possible, effectively asking the observer to pick the plot that is most different, and allowing observers to provide their own reasons for their choice. This ensures all possible deviations from the null hypothesis to be detected. However, if lineup tests are run in head-to-head comparisons with conventional statistical tests, such as the experiments in \cite{majumder:2013} or \cite{tengfei:2013}, these questions have to be phrased much more specifically, in order to avoid  type III errors \citep{mosteller:48} consisting of correctly rejecting the null hypothesis for a wrong reason. In those experiments, the structure in the data is strictly controlled in the simulation process, which allows for specific questions to be asked.

In the election example, observers were asked ``which plot is the most different?''. The type of plot, showing two (modified) stacked bar charts in different colors should suggest to the observer that the interesting feature is the difference between the two heights. Most observers based their decision on this -- besides  `asymmetry' between the towers (a reason suggested by us), most free-form responses for the reason of choice describe a comparison between the red and blue towers. In the responses for this lineup we see something that is very typical of responses in lineups: it is only a few panels from the lineup that observers pick, while most panels do not get picked or get picked only a few times. Out of a total of 72~responses, 23 observers picked panel~\#8, 12 observers picked the (data) panel \#13, 10 observers picked panel~\#4, 6 observers picked panel \#7, and all of the remaining panels were picked at most three times. From the panels picked, we see that the top three choices  are all extreme with respect to heights of blue versus red towers:  both panels \#8 and \#13 show a large difference between the height of the towers, while panel \#4 is the only panel, in which the red tower is higher than the blue one.

@@

We have conducted a series of experiments (see Table~\ref{tbl:visual_stat} for an overview)
%There have been ten experiments conducted 
using Amazon's Mechanical Turk~\citep{turk} (MTurk) for a variety to purposes: validating the protocol against existing tests, comparing plot designs, and evaluating structure in data analysis problems. In all of the experiments we collected demographic details, such as gender, age bracket, and educational background.  Mining this data will provide a basis for evaluating
%are being used to evaluate 
the effects of different covariates on an individual's ability to evaluate lineups. The design of some of the experiments allow a further investigation into aspects beyond participants' demographics: experiments~5, 6 and 7 can be used to investigate short term learning trends, while experiment~9 allows insight on whether the  positioning of the  data plot in the lineup has an effect on detection rate. 
 %Each of these experiments collected demographic details of the subjects. 
Section~\ref{sec:factor_performance} discusses human factors that we expect to be influential on the performance of the observer. Section~\ref{sec:exp_design} describes the methods used to assess the effects, and Section~\ref{sec:result_socio} summarizes the results of what we have learned about human factors affecting lineup evaluation.

@@

\begin{itemize} \itemsep 0in \parsep-10pt %\parskip0pt 
%\subsection{Choice of Visual Test Statistic}  
\item{\bf Choice of Visual Test Statistic:}  
%There are a variety of ways that
Data can be plotted in a variety of ways and we can enhance plots by axes, legends or representations of models. The choice of plot primarily should be appropriate for the problem being investigated, as the data plot is the visual test statistic. 
When for example studying the association between two quantitative variables a scatterplot is usually considered best practice. Enhancements, such as a line overlaying a scatterplot to represent a fitted model, might be appropriate for studying models for the data.  Some choices of visual test statistics provide better results than others when studying the same problem \citep{heike:2012}, because some plot types make it easier for certain structures to be seen.

Table~\ref{tbl:visual_stat} shows the large variety of visual test statistics used in the Turk studies completed so far. These experiments were conducted to examine many different visual tasks. Side-by-side boxplots were used in experiment~1 to study the importance of a categorical variable in a regression model. For both experiments~2 and~3, a scatterplot is the visual test statistic, but a regression line fitted through the points is overlaid for experiment~2. The experiments studied slightly different problems, for which the difference in plot type was important. Experiment~4 compared polar coordinates to euclidean coordinates on the reading of structure in a barchart.  Experiment~5 examined the style of plot for a bivariate data problem, by comparing results when using a scatterplot, side-by-side dot plots, overlaid densities or side-by-side boxplots. Experiment~6 studied the difference between variations of boxplots. Experiment~7 studied the effects of high dimension low sample size on the separation of groups in low-dimensional projections. The sine illusion, where people read the strength of differences between curves as different when there is a strong sinusoidal structure, is examined in experiment~8. Experiment~9 was conducted to test if there was any structure in an RNA-seq data set measured on soybeans, and used interaction plots classically constructed to examine interaction effects in $2\times 2$ factorial designs. Experiment~10 used different sorts of residual plots for hierarchical linear models as the visual test statistics.

In experiments~4,~5 and~6, different visual test statistics were used on the same data, providing the ability to examine the effect of these choices, the use of lineup tests for comparisons and  examples are discussed in \citet{heike:2012, qqplots:2015}

%\subsection{Signal in the Data}
\item{\bf Signal in the Data:}
A visual test statistic is chosen for a particular task. If it is well-designed  departures from the null hypothesis will be visible in the plot, especially if the signal is strong. An important factor that helps an observer to identify the actual data plot in a lineup is the strength of the signal in the data. Simulated data was generated for all of the experiments, except exps.~4 and~9, enabling the study of signal strength and actual data plot detection. In experiment~4 we took a different approach by varying the sampling size, which also allowed us to investigate the effect a design had on detection rate while varying signal/noise ratio. Primarily the results were as expected, that as the signal strength increased observers more frequently picked the actual data plot as different from the null plots.

\item{\bf Question Design:} A primary purpose of visual inference is to preclude the necessity to pre-specify discoverable features in data that is required by classical statistical inference. Visual inference supports discovery of structure, enabling the unexpected to surprise us.  This is tightly connected to the task that the human observer is asked to perform. For most purposes the observer is asked a very general question, such as to pick the plot that is most different from the others, and explain the reasons why they see it as different. By being general, all possible discoverable structures are included in the significance calculations \citep[see also ][]{buja:2009}.

In the initial Turk experiments (1,2,3) conducted with simulated data to compare the results from visual inference with those obtained by classical inference, the questions posed to the observers were more targeted. This was important to allow a direct comparison of the results to those from conventional testing. %in order to make a comparison of the results. 
For latter studies (4, 7, 10), questions posed to the observer were more general. Table~\ref{tbl:visual_stat} lists the questions used for each experiment.

%\subsection{Demographics} 
\item{\bf Demographics:}  During each of the experiments, data on age, gender, education level and geographic location was collected. Each observer self-reported gender, age in roughly five year intervals, education level as high school or less, some college courses, an undergraduate degree, some graduate courses or a graduate degree. The IP address of the computer afforded the geographical location of the subject. The purpose of collecting this information with each experiment is to examine the effect that they have on the results of visual inference -- ideally very little.

@@

%\subsection{Location of Actual Data Plot in the Lineup}
\item{\bf Location of Actual Data Plot in the Lineup} For all of the lineups used in the experiments a $5\times 4$ grid of 20 panels is used. A random number generator is used to determine the position where the actual data plot is placed in each lineup. Eye tracking experiments \citep{zhao:2012} suggest that some observers traverse lineups in horizontal direction, while others have a vertical up-and-down approach to their search. Almost universally, observers start at the top left of a lineup. This leads to some concern that observers might be able to more easily identify the actual data plot, if it is placed at the top left of a lineup than when it the actual data is placed at the bottom right.
%With the help of eye-tracking equipment \citet{zhao:2012} tracked the observers' eyes to see their approach in searching the plots in a lineup. The results suggest that many people read lineups from left to right, some read from upward to downward. 
%There is some concern that if the actual data plot is located in the top left of the lineup, observers may be able to identify it more easily than when the actual data plot is at the bottom right. 
Ideally, this does not happen. Experiment~9  is designed to allow us to investigate the effects of location on detection. Five different locations in the lineup were used to assess how fast and accurately observers identified the actual data plot.

\item{\bf Sample of Null Plots:} In classical inference, the test statistic is compared to the full null distribution, to decide if it is extreme or not. In visual inference the actual plot is compared to plots of a finite number of samples from the null distribution. In the lineups used in the Turk experiments, the actual data plot is compared on 19 null plots. These null samples are random draws, and there is a chance that one or more null plots in the lineup may be similar or even more extreme than the actual data plot. The sample of null plots can affect the observer's decision. This is tested with the data from experiment~9, where the experiment was set up with lineups made from different null plots.

%\subsection{Individual Skill or Ability} 
\item{\bf Individual Skill or Ability:} Each person may have different  aptitude for reading statistical plots, and their visual skill sets might be differently developed.  We can examine the effect of individuals' skills and abilities because multiple subjects evaluated the same lineups, which allows us to estimate, if some subjects consistently detect the actual data plot more often than others.  
 %For example, in a controlled experimental study \citet{zhao:2012}  noticed that some people spent a lot of time to decide no matter whether the lineup is difficult or easy while some simply glanced at lineups to make a decision. This influences the response of the observer. Also subject specific variation in the power of visual test is observed in \cite{majumder:2013}. 
\end{itemize}

@@

Two of the factors, signal in the data and individual abilities, were studied in \cite{majumder:2013}. The choice of visual test statistic was examined in \cite{heike:2012}. In each of these analyses demographic factors were given a cursory glance, to ensure that they did not have large effects on the results. The design of experiments 5, 6 and 7 enables the examination of learning trend, which is studied in this paper. Experiment~9 was a real test case for visual statistical inference, and in order to understand the significance of the structure in the genomic data, multiple lineups were made in which location of the actual data plot, and the sample of nulls, were randomized. This enables the assessment of the effect of these factors on the results. This section describes the experimental methods used to examine the effects of demography, placement of the actual data plot, sample of null plots and the existence of a learning trend.

@@

\subsection{Location Effect} \label{sec:location_design} Experiment~9 studied significant expression in an RNA-seq study, and was designed so that location effect of the actual data plot in a lineup could also be assessed. The data used, documented in \cite{atwood:2013}, measures gene expression of soybean by RNA-seq methods. Two factors were of primary interest a main effect for genotype and an interaction effect between genotype and treatment condition.

In large studies such as this there is a valid question whether the data exhibits any structure at all, or if the small $p$-values are simply occurring by chance, from the massive multiple testing. This overall significance is studied using visual inference  in \citet{tengfei:2013}.

In order to study the effect that  location has on evaluating lineups, we used multiple lineups for each of the data plots. For each data plot, five sets of null plots were generated and the actual plot was randomly placed in one of five different locations in a lineup of size 20. 
For the genotype effect, the locations were  1, 8, 12, 17, 20 and locations 2, 9, 12, 16, 20 were used for the interaction effect. %Five different sets of null plots were used to produce 5 lineups for each location position, creating 
Overall this created a total of 25 lineups for studying the genotype effect, and another 25 lineups for studying the interaction effect. Each observer saw three lineups, one for genotype, one for interaction, and one easy lineup that was used to help clean the data.

To examine if the difference in detection rate among the locations is statistically significant a one-way multivariate analysis of variance (MANOVA) model is fit to the data.
Let $\mathbf{Y}=(Y_1,Y_2, ... , Y_p)^\top$ be a vector of random variables with dimension $p = 5$, the total number of null sets, and let $\mathbf{Y}_{ij}$ represent the $j$th vector response for  location $i$ with $i=1,2, ..., I$ for $U=5$. Because the same data plot is shown in each lineup, it is assumed that there could be some association between the responses for each null set, which suggests the MANOVA model rather than a univariate ANOVA.  The MANOVA model 
\begin{equation}\label{manova}
\mathbf{Y}_{ij} = \mu_{i} + \epsilon_{ij}
\end{equation}
where $\mu_{i}= (\mu_{1i},\mu_{2i}, ..., \mu_{pi})^\top$ is the mean vector for location $i$ and $Var(\epsilon_{ij})=\Sigma$, tests for significant difference between the means.

@@

\subsection{Overview of the Data} A total of 2340 participants provided feedback data on the lineups in ten different experimental studies. Figure~\ref{fig:turker_location} displays the locations of participants around the world.  Most of the participants were from the United States and India. There were 70 other different countries represented. This provides a diverse pool of participants. The diversity in not only  geographic but also in gender, age group and education level as can be seen in Table~\ref{tbl:demographics}. The large number of female participants from all countries was a pleasant surprise to us.

@@

Besides the United States and India, countries such as Canada, Romania, the United Kingdom and Macedonia have more than 10 participants each. The remaining 64 countries have fewer than 10 participants each. The distribution of participants is similar in all ten experiments.

\begin{table}[hbtp]
\caption{Demographic information of the subjects participating in the MTurk experiments. Average time taken for evaluating a lineup is shown in seconds.}
\centering
\scalebox{.85}{
\begin{tabular}{rlrrcr}
  \hline
  && \multicolumn{2}{c}{Participants}&Average& Number of\\
  \cline{3-4}
 Factor & Levels & Total &\% & Time & Responses \\ 
  \hline
Gender & Female & 989 & 42.26 & 43.69 & 10538 \\ 
   & Male & 1351 & 57.74 & 48.75 & 13542 \\  
\hline
  Education & High school or less & 194 & 8.29 & 37.13 & 2253 \\ 
   & Under grad courses & 418 & 17.86 & 42.85 & 4068 \\ 
   & Under grad degree & 585 & 25.00 & 44.53 & 5792 \\ 
   & Graduate courses & 247 & 10.56 & 44.63 & 2471 \\ 
   & Graduate degree & 899 & 38.42 & 52.07 & 9496 \\ 
\hline
  Age & 18-25 & 743 & 31.75 & 43.50 & 7340 \\ 
   & 26-30 & 549 & 23.46 & 46.19 & 5610 \\ 
   & 31-35 & 375 & 16.03 & 44.19 & 3912 \\ 
   & 36-40 & 257 & 10.98 & 54.97 & 2714 \\ 
   & 41-45 & 140 & 5.98 & 43.89 & 1510 \\ 
   & 46-50 &  94 & 4.02 & 49.54 & 994 \\ 
   & above 50 & 184 & 7.86 & 52.32 & 2000 \\ 
\hline   
  Country & United States & 1086 & 46.41 & 39.65 & 10763 \\ 
   & India & 980 & 41.88 & 52.63 & 10238 \\ 
   & Rest of the world & 279 & 11.92 & 50.38 & 3079 \\

@@

% 
% % latex table generated in R 3.4.1 by xtable 1.8-2 package
% % Wed Oct 18 16:56:32 2017
% \begin{table}[ht]
% \centering
% \begin{tabular}{llrrrr}
%   \hline
% var & lbls & subsjects & percents & avg\_time & response \\ 
%   \hline
% gender\_level & Female & 989 & 42.26 & 43.69 & 10538 \\ 
%   gender\_level & Male & 1351 & 57.74 & 48.75 & 13542 \\ 
%   degree & High school or less & 194 & 8.29 & 37.13 & 2253 \\ 
%   degree & Under grad courses & 418 & 17.86 & 42.85 & 4068 \\ 
%   degree & Under grad degree & 585 & 25.00 & 44.53 & 5792 \\ 
%   degree & Graduate courses & 247 & 10.56 & 44.63 & 2471 \\ 
%   degree & Graduate degree & 899 & 38.42 & 52.07 & 9496 \\ 
%   age\_level & 18-25 & 743 & 31.75 & 43.50 & 7340 \\ 
%   age\_level & 26-30 & 549 & 23.46 & 46.19 & 5610 \\ 
%   age\_level & 31-35 & 375 & 16.03 & 44.19 & 3912 \\ 
%   age\_level & 36-40 & 257 & 10.98 & 54.97 & 2714 \\ 
%   age\_level & 41-45 & 140 & 5.98 & 43.89 & 1510 \\ 
%   age\_level & 46-50 &  94 & 4.02 & 49.54 & 994 \\ 
%   age\_level & above 50 & 184 & 7.86 & 52.32 & 2000 \\ 
%   country & United States & 1086 & 46.41 & 39.65 & 10763 \\ 
%   country & India & 980 & 41.88 & 52.63 & 10238 \\ 
%   country & Rest of the world & 279 & 11.92 & 50.38 & 3079 \\ 
%    \hline
% \end{tabular}
% \end{table}

The largest number of participants falls within the age group of 18 to 25, with the majority being between 18 to 35. Many older participants also took part in the experiments. %Interestingly, there are many participants in the older age groups as well. 
The United States shows a more uniform participation beyond age 30,  participants from India are primarily younger.

In terms of academic background, the majority of participants had a graduate degree (a total of 899 participants, about 38.42\%). Examining the participants' location in conjunction with the academic background reveals that there is an above average  number of participants from India reporting to have a graduate degree. In retrospect this turned out to be a culturally based misunderstanding, as any university degree in India is considered to be a graduate degree, while the North American definition of a graduate degree refers to a Masters level education or above. Most of the  participants from the U.S. report to have an undergraduate degree or at least have some undergraduate courses.

Compared to data from the 2010 US Census \cite{census}, participants were both relatively younger, and more highly educated than the general US population. 
Being turkers implicitly assures that the participants in these experiments have a relatively high level of technological fluency. In fact, the demographics of participants align somewhat better with demographics of internet users published by
\citet{pew}.

The distribution of male and female participants are similar among all age groups except for age group 18-25 in India where the proportion of female participants is lower. The distribution of education levels also differs slightly across countries for this age group.

A total of 1912 lineups were evaluated in the ten experiments. Each person evaluated at least 10 lineups except for experiment~9 where each participant evaluated three lineups.
As a measure to ensure a high quality in the evaluations,
 a test plot was shown to each observer, and the performance on this test plot was used for screening purposes. %A test plot was shown to each observer and the feedback received from that plot is used to ensure data quality. 
Test plots were chosen in a way that any serious attempt at answering the lineup would result in a correct answer.
Data from participants who gave the wrong answer on the test plot % were considered as gaming the system, and their data were removed.
led to the removal of all of their answers from the analysis. 
Results from test plots were also not considered in any of the analyses.
 In some cases  participants chose to not provide their demographic information. Also, for some ip address, the actual geographical locations could not be retrieved. This resulted in some missing geographic information.

To explore the significance of the demographic factors model~\eqref{eqn:demographic_response} is fit to the data, with age, country, education and gender as fixed effects. To estimate the significance of each of the factors, reduced models are fit with that factor removed from the model. Table~\ref{tbl:anova_factor} summarizes the results. Except for gender, all of the demographic factors show significant differences in  detection rates.  With respect to time taken, all of the demographic factors are significant.

>>>>

The lineup protocol introduced by \citet{buja:2009} quantifies the significance of graphical findings in exploratory data analysis, forming part of visual statistical inference. \citet{majumder:2013} extended and validated this approach through simulation studies, showing that visual inference can sometimes outperform conventional methods. They also defined the power of visual tests and proposed a way to calculate it for a given lineup. In visual inference, the test statistic is a plot of observed data, randomly placed among null plots generated under a null hypothesis, often assuming no structure. Observers evaluate the lineup, and if they consistently identify the data plot, this provides evidence against the null hypothesis, akin to a conventional hypothesis test.

A lineup can be evaluated by one or more observers, and a distribution similar to the binomial, adjusted for dependencies from the lineup scenario, is used to calculate the $p$-value based on how often the actual data plot is identified \citep{vinference}. To avoid expectation bias \citep{meilgaard}, observers must not know the data constituting the lineup or have seen the actual data plot beforehand. The accompanying question in a lineup should be broad, asking observers to pick the most distinct plot, allowing detection of any deviations from the null hypothesis. However, in head-to-head tests with conventional methods \citep{majumder:2013, tengfei:2013}, specific questions are needed to avoid type III errors \citep{mosteller:48} consisting of correctly rejecting the null hypothesis for a wrong reason.

@@

We conducted several experiments (see Table~\ref{tbl:visual_stat} in the Appendix~\ref{sec:tbl_fig}) using Amazon's Mechanical Turk~\citep{turk} (MTurk) for validating the lineup protocol, comparing plot designs, and evaluating data analysis structures. Demographic data such as gender, age, and educational background were collected to assess the influence of these covariates on lineup evaluations. Additionally, experiments investigated the short-term learning trends.

Section~\ref{sec:factor_performance} discusses expected human factors influencing observer performance and Section~\ref{sec:result_socio} presents findings on factors affecting lineup evaluations. The methods used to assess the effects are oulined in Appendix~\ref{sec:exp_design}.

@@

\begin{itemize} \itemsep 0in \parsep-10pt %\parskip0pt

\item{\bf Demographics:}  During each of the experiments, data on age, gender, education level and geographic location was collected. Each observer self-reported gender, age in roughly five year intervals, education level as high school or less, some college courses, an undergraduate degree, some graduate courses or a graduate degree. The IP address of the computer afforded the geographical location of the subject.

@@

%\subsection{Individual Skill or Ability} 
\item{\bf Individual Skill or Ability:} Each person may have different  aptitude for reading statistical plots, and their visual skill sets might be differently developed.  We can examine the effect of individuals' skills and abilities because multiple subjects evaluated the same lineups, which allows us to estimate, if some subjects consistently detect the actual data plot more often than others.

\end{itemize}

 \subsection{Overview of the Data} In ten experimental studies, a total of 2,340 participants from 70 countries, primarily the United States and India, provided feedback data on lineups, ensuring a diverse pool of respondents in terms of geography, gender, age, and education. There were a high number of female participants(42\%) across most countries, which was an unexpected but positive outcome(see Table~\ref{tbl:turker_location} in Appendix~\ref{sec:tbl_fig}). Notably, countries like Canada, Romania, the United Kingdom, and Macedonia also had notable participant representation.

The majority of participants were aged 18 to 35, with the United States showing a more balanced age distribution compared to India, where younger individuals were more prevalent. Educational backgrounds were diverse, with 38.42\% of participants reporting a graduate degree. However, there was a cultural misunderstanding as Indian participants classified any university degree as a graduate degree, while the North American definition refers to a Master’s degree or higher. Compared to the 2010 US Census \cite{census}, the participants skewed younger and more highly educated.

A total of 1,912 lineups were evaluated across the studies, with each participant reviewing at least 10 lineups, except for experiment~9, where they reviewed three. To ensure quality, participants were given test plots to evaluate first, and incorrect responses led to their data being excluded from the analysis. Some demographic and geographical data were missing due to participants opting out or issues retrieving IP-based location data.

To explore the significance of demographic factors in detecting the data plot in the lineup, a mixed-effects model~\eqref{eqn:demographic_response} is fitted to the data, with age, country, education, and gender as fixed effects. To estimate the significance of each of the factors, reduced models are fitted with that factor removed from the model. Table~\ref{tbl:model_result_demographics} summarizes the results. Except for gender, all of the demographic factors show significant differences in  detection rates.  With respect to time taken, all of the demographic factors are significant.

The first level of each factor serves as the baseline for the model. The age group 36-40  has a significantly higher detection rate than the 18-25 age group, and to a lesser extent this is also true for other age groups except for 26-30. Participants from the rest of the world have a similar detection rate to those from the USA, but participants from India show a significantly lower detection rate. Subjects reporting to have  an undergraduate degree had a significantly higher detection rate. For time to respond, all age groups were significantly faster than  18-25 year olds.  Participants from India as well as the rest of the world were faster than US Americans. A higher degree corresponded to longer time, and males were on average faster than females.

Figure~\ref{fig:demographic_effect} supports the model results. The average time taken to respond (natural log) and detection rate are computed by lineup for the different demographic factor levels, and displayed using boxplots. Mean values are shown as dots inside the boxplot, as this statistic rather than the median is used in the models. The means and medians for response time (top row) are essentially equal. There are some differences between the two in the detection rate plots (bottom row). The differences in these values for different factors, that was tested by the models, can be seen in the plots. However, particularly for detection rate, the variation is huge. A large component of the variation is difficulty of the lineup. In some lineups the actual data plot was distinctively different from the null plots, and we would expect that the detection rate would be close to 100\%. In other lineups there was no difference, making it very difficult to evaluate, with low detection rate and longer time to respond.

In Table~\ref{tbl:model_result_demographics} notice that lineup specific error is estimated as 2.27 which is much higher than any of the other parameter estimates in the model. This indicates that the major and most important factor affecting the detection rate is  lineup difficulty.  So although the demographic factors emerge from the model to be statistically significant, the practical significance is minimal. We illustrate this with the following example of graduate degree.

While some of the demographic factors are strongly statistically significant, the main source of variation in detection rate is the lineup difficulty. For example, let's examine the effect of an undergraduate degree. To see just how large the effect is, we examine the change in detection rate for a (hypothetical) 18-25 year old female in the United States, with some graduate course work as compared to an undergraduate degree, for an average difficulty lineup (random effect of zero). Plugging in the relevant quantities to the fitted model gives a difference equal to:
$$ \frac {\exp(-0.63-0.13)}{1+\exp(-0.63-0.13)}-  \frac {\exp(-0.63+0.16)}{1+\exp(-0.63+0.16)} = 0.319- 0.385 =-0.066.$$

The person with some graduate course work picks the data plot  on average in 31.9\% of lineups of average difficulty, as compared to 38.5\% if they have an undergraduate degree. This difference is reduced to 4\% for a lineup with one standard deviation order of magnitude difference in difficulty. For two standard deviations it further reduces to 0.6\%. Thus, although there is a statistically significant difference in the proportion at which participants identify the  data plot for some demographic factors, these are not  significant in any practical sense.  Figure~\ref{practical_impact_und_und_graduate.pdf} illustrates this example showing fitted models for a US 18-25 female with either a high school education or an undergraduate degree.  Similar calculations show the same negligible impact of age level 36-40 (\Sexpr{round(exp(-0.63+0.29)/(1+ exp(-0.63+0.29)) - exp(-0.63)/(1 + exp(-0.63)),3)} at most)  and country India (\Sexpr{round(exp(-0.63-0.19)/(1+ exp(-0.63-0.19)) - exp(-0.63)/(1 + exp(-0.63)), 3)} at most) on the probability of a correct response. Thus even though some of the demographic factors are statistically significant, practically, demographics do not substantially influence the results.

\subsection{Learning Trend} Models \eqref{eqn:trend_response} and \eqref{eqn:trend_time} are fitted to the data from experiments~5,~6, and~7 separately. Each participant evaluated 10 lineups in 10 attempts. In this model, attempt is fitted as a factor variable to allow for all possible non-linear learning trends.  It should be noted that we also examined an alternative to model~\eqref{eqn:trend_response} where attempt was linearly fitted as a continuous covariate, but this also was not significant.

The parameter estimates and $p$-values of the fixed effects from model~\eqref{eqn:trend_response} are presented in Table~\ref{tbl:model_result_response} in the Appendix~\ref{sec:exp_design}. Only for experiment~6 there is some evidence of a learning curve. There are marginally small $p$-values for most of the attempt levels, and positive parameter estimates, suggesting that more attempts increases detection rate.

Table~\ref{tbl:model_result_time} in the Appendix~\ref{sec:exp_design} presents the results of model~\eqref{eqn:trend_time} where response time is examined with respect to attempt. Attempt is modeled as a linear covariate here, with a shift factor used to adjust for additional time on the first lineup evaluated. Interestingly, time to respond significantly decreases as number of attempts increases, for all three experiments. 
The parameter $\alpha$ for fixed effect covariate attempt is highly significant in all the experiments. The negative estimates suggest that on average later attempts took less time. Even though observers did not improve in their performance in successive attempts, they became more efficient in their response. The parameter $\alpha_1$ for first attempt is also highly significant. The positive estimates of $\alpha_1$ indicates that first attempt made by an observer required much more time than any of the other attempts. One explanation might be that participants take the chance to  carefully read through instructions and familiarize themselves  with the types of plots used in the experiment before evaluating their first lineup, which would inflate the time taken on the first lineup. Each of the experiments asks observers to give a reason for their choice (out of a list of pre-defined answers. For each lineup evaluation, participants are also asked to state how confident in their choice they are (on a scale of 1 to 5).  Later pages of the web site were identical  except for the lineup.

% ---------------------------------------- Appendix ----------------------------------

\appendix

@@

Two of the factors, signal in the data and individual abilities, were studied in \cite{majumder:2013}. The choice of visual test statistic was examined in \cite{heike:2012}. In each of these analyses demographic factors were given a cursory glance, to ensure that they did not have large effects on the results. The design of experiments 5, 6 and 7 enables the examination of learning trend, which is studied in this paper. Experiment~9 was a real test case for visual statistical inference, and in order to understand the significance of the structure in the genomic data, multiple lineups were made in which location of the actual data plot, and the sample of nulls, were randomized. This enables the assessment of the effect of these factors on the results. This section describes the experimental methods used to examine the effects of demography, sample of null plots and the existence of a learning trend.

@@

@@

% ===================================================================================
\section{Some tables and figures}\label{sec:tbl_fig}
%------------------------------------------------------------------------------------

@@

\begin{table}[hbtp]
\caption{\label{tbl:turker_location}Demographic information of the subjects participating in the MTurk experiments. Average time taken for evaluating a lineup is shown in seconds.}
\centering
\scalebox{.85}{
\begin{tabular}{rlrrcr}
  \hline
  && \multicolumn{2}{c}{Participants}&Average& Number of\\
  \cline{3-4}
 Factor & Levels & Total &\% & Time & Responses \\ 
  \hline
Gender & Female & 989 & 42.26 & 43.69 & 10538 \\ 
   & Male & 1351 & 57.74 & 48.75 & 13542 \\  
\hline
  Education & High school or less & 194 & 8.29 & 37.13 & 2253 \\ 
   & Under grad courses & 418 & 17.86 & 42.85 & 4068 \\ 
   & Under grad degree & 585 & 25.00 & 44.53 & 5792 \\ 
   & Graduate courses & 247 & 10.56 & 44.63 & 2471 \\ 
   & Graduate degree & 899 & 38.42 & 52.07 & 9496 \\ 
\hline
  Age & 18-25 & 743 & 31.75 & 43.50 & 7340 \\ 
   & 26-30 & 549 & 23.46 & 46.19 & 5610 \\ 
   & 31-35 & 375 & 16.03 & 44.19 & 3912 \\ 
   & 36-40 & 257 & 10.98 & 54.97 & 2714 \\ 
   & 41-45 & 140 & 5.98 & 43.89 & 1510 \\ 
   & 46-50 &  94 & 4.02 & 49.54 & 994 \\ 
   & above 50 & 184 & 7.86 & 52.32 & 2000 \\ 
\hline   
  Country & United States & 1086 & 46.41 & 39.65 & 10763 \\ 
   & India & 980 & 41.88 & 52.63 & 10238 \\ 
   & Rest of the world & 279 & 11.92 & 50.38 & 3079 \\

@@
